The quench action approach in finite integrable spin chains

### 2015-12-07 QAMC_draft/fs_qa_v8.tex
arxiv

We consider the problem of constructing the stationary state following a quantum quench, 
using the exact overlaps for finite size integrable models. 
We focus on the isotropic Heisenberg spin chain with initial state N\'eel or Majumdar-Ghosh (dimer),
although the proposed approach is valid for an arbitrary integrable model. 
We consider only those states which do not contain zero-momentum strings because the latter 
are affected by fictitious singularities that are very difficult to take into account. 
We show that the fraction of eigenstates that do not contain zero-momentum strings is vanishing in the thermodynamic limit.
Consequently, restricting to this part of the Hilbert space would lead to vanishing expectation values of local observables. 
It is, however, possible to reconstruct the asymptotic values by properly reweighting the expectations in the considered 
subspace at the price of introducing finite-size corrections. 
We also develop a Monte Carlo sampling of the Hilbert space which allows us to study larger systems. 
We accurately reconstruct the expectation values of the conserved charged and the root distributions in the stationary state, 
which turn out to match the exact thermodynamic results. 
The proposed method can be  implemented even in cases in which an analytic thermodynamic solution is not obtainable.

@@

In integrable models, however, the presence of local or 
quasi-local integrals of motion strongly affects the dynamics, preventing the 
onset of thermal behavior. It has been suggested in Ref.~\cite{rigol-2007,
rigol-2008} that in this situation the post-quench steady-state can be described 
by the so-called Generalized Gibbs Ensemble ($GGE$) as 
%
\begin{equation}
\label{gge}
\langle{\mathcal O}\rangle_{GGE}=\frac{1}{Z}\Tr({\mathcal O}\rho^{GGE}), 
\quad\textrm{with}\quad\rho^{GGE}\equiv\frac{1}{Z}\exp\Big(-\sum_j\beta_jQ_j\Big). 
\end{equation}
%
Here $Q_j$ are mutually commuting conserved quantities, i.e., $[{\mathcal H},Q_j]=0\,
\forall j$ and $[Q_j,Q_k]=0\,\forall j,k$, whereas $\beta_j$ are Lagrange multipliers 
to be fixed by imposing that  $\langle Q_j\rangle_{GGE}=\langle\Psi_0|Q_j|\Psi_0\rangle$. 
%Note that for the diagonal ensemble~\eref{d-ensemble}  one has trivially that $\langle
%{\mathcal O}\rangle_{DE}=\langle\Psi_0|{\mathcal O}|\Psi_0\rangle$. 
This also provides  useful sum rules that will be of interest in this paper. 
The validity of the GGE has been confirmed in non-interacting theories~\cite{calabrese-2006,cramer-2008,barthel-2008,rossini-2009,calabrese-2011,cazalilla-2012,se-12,mossel-2012a,collura-2013,fagotti-2013,kcc14,kcc14a,sotiriadis-2014,bkc-14}, 
whereas in interacting ones the scenario is 
still not fully settled and some recent works~\cite{ilievski-2015a,cardy-2015} suggest that the $GGE$ 
description of the steady state  is complete provided that the so-called quasi-local 
charges~\cite{prosen-2014,pereira-2014,ilievski-2015,essler-2015} are included in the GGE~\eref{gge}.

@@

\paragraph*{Outline of the results.} 
In this paper, by combining exact Bethe ansatz and Monte Carlo techniques, we investigate 
the diagonal ensemble and the quench action approach in {\it finite} size integrable models. 
%
We consider the spin-$1/2$ XXZ spin-chain with $L$ sites, which is defined by the Hamiltonian 
%
\begin{equation}
\label{xxx-ham}
{\mathcal H}\equiv J\sum\limits_{i=1}^L\left[\frac{1}{2}(S_i^+S^-_{i+1} 
+S_i^{-}S_{i+1}^+)+\Delta S_i^zS_{i+1}^z-\frac{1}{4}\right].
\end{equation}
%
Here $S^{\pm}_i\equiv (\sigma_i^x\pm i\sigma_i^y)/2$ are spin operators acting on the 
site $i$, $S_i^z\equiv\sigma_i^z/2$, and $\sigma^{x,y,z}_i$ the Pauli matrices. We 
fix $J=1$ in~\eref{xxx-ham} and use periodic boundary conditions, identifying sites 
$L+1$ and $1$. The total magnetization $S_{T}^z\equiv\sum_iS_i^z=L/2-M$, with $M$ number 
of down spins (particles), commutes with~\eref{xxx-ham}, and it is used to label its 
eigenstates. The $XXZ$ chain is Bethe ansatz solvable~\cite{bethe-1931,kor-book} and its eigenstates 
are in one-to-one correspondence with the solutions of the so-called Bethe equations 
(see~\ref{sec:1.2}). 
%
The non-equilibrium quench dynamics of the XXZ spin chain has been investigated numerically intensively \cite{dmcf-06,bar-09,krs-12,cbp-13,cce-15,fagotti-2014},
the GGE with all the local charges has been explicitly constructed \cite{fagotti-2014,pozsgay-2013,fagotti-2013b}, 
and the quench action solution provided for some initial states \cite{pozsgay-2014A,wouters-2014A}.

For concreteness reasons, we focus on the isotropic Heisenberg point ($XXX$), i.e. Hamiltonian \eref{xxx-ham} with $\Delta=1$, 
but we stress that our approach applies to arbitrary $\Delta$, and  also to arbitrary integrable models. 
We consider the quenches from the zero-momentum N\'eel 
state $|N\rangle$ and the Majumdar-Ghosh state $|MG\rangle$, which are defined as  
%
\begin{equation}
\label{psi0}
|N\rangle\equiv\Big(\frac{|N_1\rangle
+|N_2\rangle}{\sqrt{2}}\Big), \qquad 
|MG\rangle\equiv \Big(\frac{|\uparrow\downarrow\rangle-|\downarrow\uparrow\rangle}
{\sqrt{2}}\Big)^{\otimes L/2}. 
\end{equation}
%
Here $|N_1\rangle\equiv |\uparrow\downarrow\rangle^{\otimes L/2}$, and $|N_2\rangle
\equiv |\downarrow\uparrow\rangle^{\otimes L/2}$, with $\otimes$ denoting the tensor 
product. Note that both $|N\rangle$ and $|MG\rangle$ are invariant under one site 
translations. Our study relies on the analytical knowledge of the overlaps between 
the N\'eel and Majumdar-Ghosh state and the eigenstates of~\eref{xxx-ham}~\cite{
brockmann-2014,brockmann-2014b,brockmann-2014c,pozsgay-2014a,piroli-2014,mazza-2015}.

@@

%The key idea is to sample the eigenstates with no zero-momentum 
%strings with the diagonal ensemble probability measure $\rho^{DE}$ (cf.~\eref{de1}). 
We also develop a Monte Carlo scheme, which is a generalization of 
the approach presented in Ref.~\cite{alba-2015} to simulate the Generalize 
Gibbs Ensemble. The method is based on the knowledge of the N\'eel overlaps 
$\langle\lambda|N\rangle$ and on the knowledge of the Hilbert space structure 
of the $XXX$ chain in the Bethe ansatz formalism. The approach allows us to 
simulate effectively chain with $L\lesssim 60$, although larger systems 
sizes can in principle be reached.

Strikingly, after the reweighting, although for small chains violations of the conserved quantities 
sum rules are present, these violations vanish in the thermodynamic limit 
and the sum rules are restored. This implies that the only effect of the 
Hilbert space truncation is to introduce finite size scaling corrections. 
In the quench action language, this means that the eigenstates corresponding to non 
zero-momentum strings contain enough physical information about saddle point. 
This is numerically confirmed by extracting the so-called saddle point 
root distributions, which in the Bethe ansatz language fully characterize 
the diagonal ensemble averages in the thermodynamic limit. In the numerical  
 approach these are obtained from the histograms of the Bethe ansatz 
solutions sampled during the Monte Carlo. A part from finite size scaling 
corrections, we observe striking agreement with analytical results, at least 
for the first few root distributions.

%It is useful to rewrite~\eref{d-ensemble} as 
%
%\begin{equation}
%\label{qa-prob}
%\langle{\mathcal O}\rangle=\sum\limits_{\alpha}\rho^{DE}(\alpha)
%\langle\alpha|{\mathcal O}|\alpha\rangle,\quad\textrm{with}\quad 
%\rho^{DE}(\alpha)=\exp(-2\Re{\mathcal E}(\alpha)).
%\end{equation}
%
%Here $\rho^{DE}(\alpha)$ is a diagonal ensemble density matrix, and ${\mathcal E}(\alpha)
%\equiv-\ln\langle\alpha|\Psi_0\rangle$, with $\Re$ denoting the real part.

%The N\'eel state is defined as 
%
%\begin{equation}
%\label{N\'eel}
%|N\rangle\equiv\frac{1}{\sqrt{2}}\big(|N_1\rangle
%+|N_2\rangle\big), 
%\end{equation}
%

%The Majumdar-Ghosh state $|MG\rangle$ is defined as 
%
%\begin{equation}
%\label{mg}
%|MG\rangle\equiv \Big(\frac{|\uparrow\downarrow\rangle-|\downarrow\uparrow\rangle}
%{\sqrt{2}}\Big)^{\otimes L/2}. 
%\end{equation}
%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{The spin-$\frac{1}{2}$ Heisenberg chain}
%\label{sec:1.1}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Counting and string content of the eigenstates with nonzero N\'eel and 
%Majumdar-Ghosh overlap}
\subsection{Counting and string content of parity-invariant eigenstates}
\label{app-1}

Here we provide some exact combinatorial formulas for the total number of 
parity-invariant eigenstates of the $XXX$ chain, which are the only eigenstates 
having, in principle, non-zero overlap with the N\'eel state and the 
Majumdar-Ghosh state~\cite{brockmann-2014}. 
Parity-invariant eigenstates correspond to solutions of the Bethe equations 
containing only pairs of rapidities with opposite sign. In turn, these  
eigenstates are in one-to-one correspondence with parity-invariant BGT quantum number 
configurations. %In~\ref{app-1.1} we consider all possible parity-invariant 
%eigenstates, whereas in~\ref{app-1.2} we exclude the eigenstates containing 
%zero-momentum strings.

@@

We anticipate that the total number of parity-invariant eigenstates $Z_{Neel}$ 
for a chain of length $L$ is given as 
%
\begin{equation}
\label{N-count}
Z_{Neel}=2^{\frac{L}{2}-1}+\frac{1}{2}B\Big(\frac{L}{2},\frac{L}{4}\Big)+1. 
\end{equation}
%
On the other hand, after excluding the zero-momentum strings one obtains 
%
\begin{equation}
\label{N-count-nz}
\widetilde Z_{Neel}=B\Big(\frac{L}{2},\frac{L}{4}\Big). 
\end{equation}
%
Note that~\eref{N-count} is only an upper bound for the number of eigenstates 
with non-zero N\'eel overlap, while~\eref{N-count-nz} is exact. Before 
proceeding, we should stress that the total number of particles  
in the eigenstates having non-zero overlap with the N\'eel and Majumdar-Ghosh 
states is fixed to $L/2$, due to the fact that for both states one has $S_T^z=0$. 
However, since the N\'eel state is not $SU(2)$ invariant, eigenstates with 
non-zero N\'eel overlap can contain infinity rapidities. Thus, one has to 
consider all the possible sectors with $L/2=\ell+N_{\infty}$, and $\ell$ 
($N_\infty$) the number of finite (infinite) rapidities. Notice that this 
is different for the Majumdar-Ghosh state, for which only the parity-invariant 
eigenstates in the sector with $\ell=L/2$ have to be considered (see below).

@@

Let us first consider the eigenstate sector with fixed number of finite rapidities 
$\ell$, the remaining $L/2-\ell$ ones being infinite (see section~\ref{sec:1.2}). 
Let us denote the associate string content as ${\mathcal S}=\{s_1,s_2,\dots,
s_{\ell}\}$. Here $s_n$ is the number of $n$-strings, with the constraint 
$\sum_k ks_k=\ell$. 
It is straightforward to check that 
the total number of parity-invariant quantum number pairs ${\mathcal N}_n(L,{
\mathcal S})$ in the $n$-string sector is given as 
%
\begin{equation}
\label{NnLS}
{\mathcal N}_n(L,{\mathcal S})=\Big\lfloor\frac{L}{2}-\frac{1}{2}
\sum_{m=1}^{\ell}t_{nm}s_m\Big\rfloor,
\end{equation}
%
where $t_{nm}\equiv 2\textrm{Min}(n,m)-\delta_{n,m}$. Thus, the number of parity-invariant 
eigenstates of the $XXX$ chain ${\mathcal N}(L,{\mathcal S})$ compatible with string content 
${\mathcal S}$ is obtained by choosing in all the possible ways the associated parity-invariant 
quantum number pairs as     
%
\begin{equation}
\label{NLS}
{\mathcal N}(L,{\mathcal S})=\prod_{m=1}^{\ell} B\left({\mathcal N}_m,\left\lfloor
\frac{s_m}{2}\right\rfloor\right).
\end{equation}
%
Here the product is because each string sector is treated independently, while the 
factor $1/2$ in $s_m/2$ is because since all quantum numbers are organized in pairs, 
only half of them have to be specified. Note that in each $n$-string sector only one 
zero momentum (i.e., zero quantum number) string is allowed, due to the fact that 
repeated solutions of the BGT equation are discarded. Moreover, from~\eref{NnLS} one has 
that $s_m$ is odd (even) only if this zero momentum string is (not) present.

We now proceed to consider the string configurations with fixed particle number $\ell\le 
L/2$ and fixed number of strings $1\le q\le\ell$. Note that due to parity invariance 
$\ell$ must be even. 
Also, in determining $q$ strings of different length are treated equally, i.e., 
$q=\sum_m s_m$. For a given fixed pair 
$\ell,q$ the total number of allowed quantum number configurations by definition is given 
as 
%
\begin{equation}
\label{NLlq}
{\mathcal N}'(L,\ell,q)=\sum\limits_{\{\{s_m\}\,:\, \sum m s_m=\ell, \sum s_m=q\}}
{\mathcal N}(L,{\mathcal S}),
\end{equation}
%
where the sum is over the string content $\{s_m\}_{m=1}^\ell$ compatible with the constraints 
$\sum_m s_m=q$ and $\sum_m m s_m=\ell$. The strategy now is to write a recursive relation 
in both $\ell,q$ for ${\mathcal N}'(L,\ell,q)$. It is useful to consider a shifted string 
content ${\mathcal S}'$ defined as  
%
\begin{equation}
{\mathcal S}'\equiv \{s_{m+1}\}\quad\textrm{with}\, s_m\in{\mathcal S},\,\forall m.
\end{equation}
%
Using the definition of $t_{ij}$, it is straightforward to derive that  
%
\begin{equation}
t_{ij}=t_{i-1,j-1}+2,
\end{equation}
%
which implies that ${\mathcal N}_n(L,{\mathcal S})$ (see~\eref{NnLS}) satisfies the 
recursive equation 
%
\begin{equation}
\label{inter}
{\mathcal N}_n(L,{\mathcal S})={\mathcal N}_{n-1}(L-2q,{\mathcal S}'). 
\end{equation}
%
After substituting~\eref{inter} in~\eref{NLS} one obtains 
%
\begin{equation}
\label{NLSr}
{\mathcal N}(L,{\mathcal S})=B\Big({\mathcal N}_1(L,{\mathcal S}),\Big\lfloor 
\frac{s_1}{2}\Big\rfloor\Big){\mathcal N}(L-2q,{\mathcal S}'). 
\end{equation}
%
Finally, using~\eref{NLSr} in~\eref{NLlq}, one obtains a recursive relation for 
${\mathcal N}'(L,\ell,q)$ as 
%
\begin{equation}
\label{NpLlq}
{\mathcal N}'(L,\ell,q)=\sum_{s=0}^{q-1}B\Big(\frac{L}{2}-q+\Big\lfloor
\frac{s}{2}\Big\rfloor,\left\lfloor\frac{s}{2}\right\rfloor\Big){\mathcal N}'
\left(L-2q,\ell-q,q-s\right), 
\end{equation}
%
with the condition that for $\ell=q$ one has 
%
\begin{equation}
{\mathcal N}'(L,q,q)=B\Big(\Big\lfloor\frac{L-q}{2}\Big\rfloor,\Big\lfloor
\frac{q}{2} \Big\rfloor\Big).
\end{equation}
%
This is obtained by observing that if $\ell=q$ only $1$-strings are allowed 
and~\eref{NnLS} gives ${\mathcal N}_n(L,{\mathcal S})=\lfloor (L-q)/2\rfloor$. 
It is straightforward to check that for $q$ even the ansatz 
%
\begin{equation}
\label{inter1}
{\mathcal N}'(L,\ell,q)=\frac{q}{\ell}B\Big(\frac{L-\ell}{2},\frac{q}{2}\Big)
B\Big(\frac{\ell}{2},\frac{q}{2}\Big),
\end{equation}
% 
satisfies~\eref{NpLlq}. Instead, for $q$ odd the solution of~\eref{NpLlq} is 
%
\begin{equation}
\label{inter2}
{\mathcal N}'(L,\ell,q)=\frac{\ell-q+1}{\ell}B\Big(\frac{L-\ell}{2},\frac{q-1}{2}
\Big)B\Big(\frac{\ell}{2},\frac{q-1}{2}\Big).
\end{equation}
%
The number of eigenstates in the sector with $\ell$ particles having nonzero N\'eel 
overlap $Z'_{Neel}(L,\ell)$ is obtained from~\eref{inter1} and~\eref{inter2} by 
summing over all possible values of $q$ as 
%
\begin{equation}
\label{sum1}
Z'_{Neel}(L,\ell)=\sum\limits_{q=1}^\ell {\mathcal N}'(L,\ell,q).
\end{equation}
%
It is convenient to split the summation in~\eref{sum1} considering odd and even 
$q$ separately. For $q$ odd one obtains 
%
\begin{equation}
\sum\limits_{k=0}^{\ell/2-1} {\mathcal N}'(L,\ell,2k+1)=B\Big(\frac{L}{2}-1,
\frac{\ell}{2}-1\Big),
\end{equation}
%
while for $q$ even one has 
%
\begin{equation}
\sum\limits_{k=0}^{\ell/2} {\mathcal N}'(L,\ell,2k)=B\Big(\frac{L}{2}-1,
\frac{\ell}{2}\Big). 
\end{equation}
%
Putting everything together one obtains 
%
\begin{equation}
\label{N-count-p}
Z'_{Neel}(L,\ell)=B\Big(\frac{L}{2}-1,
\frac{\ell}{2}-1\Big)+B\Big(\frac{L}{2}-1,
\frac{\ell}{2}\Big). 
\end{equation}
%
The total number of eigenstates with nonzero N\'eel overlap $Z_{Neel}(L)$ 
(cf.~\eref{N-count}) is obtained from~\eref{N-count-p} by summing over the allowed 
values of $\ell=2k$ with $k=0,1,\dots,L/2$. Note that the sum is over $\ell$ even 
due to the parity invariance.

@@

\bibitem{ga-15}
G. Goldstein and N. Andrei, % Equilibration and generalized Gibbs ensemble for hard wall boundary conditions,
Phys. Rev. B \textbf{92}, 155103 (2015).

>>>>

We consider the problem of constructing the stationary state following a quantum quench, 
using the exact overlaps for finite size integrable models. 
We focus on the isotropic Heisenberg spin chain with initial state N\'eel or Majumdar-Ghosh (dimer),
although the proposed approach is valid for an arbitrary integrable model. 
We consider only eigenstates which do not contain zero-momentum strings because the latter 
are affected by fictitious singularities that are very difficult to take into account. 
We show that the fraction of eigenstates that do not contain zero-momentum strings is vanishing 
in the thermodynamic limit. Consequently, restricting to this part of the Hilbert space leads to 
vanishing expectation values of local observables. However, it is possible to reconstruct the 
asymptotic values by properly reweighting the expectations in the considered subspace, at the 
price of introducing finite-size corrections. We also develop a Monte Carlo sampling of the 
Hilbert space which allows us to study larger systems. We accurately reconstruct the expectation 
values of the conserved charged and the root distributions in the stationary state, which turn 
out to match the exact thermodynamic results. The proposed method can be implemented even in 
cases in which an analytic thermodynamic solution is not obtainable.

@@

In integrable models, however, the presence of local or 
quasi-local integrals of motion strongly affects the dynamics, preventing the 
onset of thermal behavior. It has been suggested in Ref.~\cite{rigol-2007,
rigol-2008} that in this situation the post-quench steady-state can be described 
by the so-called Generalized Gibbs Ensemble ($GGE$) as 
%
\begin{equation}
\label{gge}
\langle{\mathcal O}\rangle_{GGE}=\frac{1}{Z}\Tr({\mathcal O}\rho^{GGE}), 
\quad\textrm{with}\quad\rho^{GGE}\equiv\frac{1}{Z}\exp\Big(-\sum_j\beta_jQ_j\Big). 
\end{equation}
%
Here $Q_j$ are mutually commuting conserved quantities, i.e., $[{\mathcal H},Q_j]=0\,
\forall j$ and $[Q_j,Q_k]=0\,\forall j,k$, whereas $\beta_j$ are Lagrange multipliers 
to be fixed by imposing that  $\langle Q_j\rangle_{GGE}=\langle\Psi_0|Q_j|\Psi_0\rangle$. 
This also provides  useful sum rules that will be of interest in this paper. 
The validity of the GGE has been confirmed in non-interacting theories~\cite{calabrese-2006,
cramer-2008,barthel-2008,rossini-2009,calabrese-2011,cazalilla-2012,se-12,mossel-2012a,
collura-2013,fagotti-2013,kcc14,kcc14a,sotiriadis-2014,bkc-14}, whereas in interacting 
ones the scenario is still not fully settled and some recent works~\cite{ilievski-2015a,
cardy-2015} suggest that the $GGE$ description of the steady state  is complete provided 
that the so-called quasi-local charges~\cite{prosen-2014,pereira-2014,ilievski-2015,essler-2015} 
are included in the GGE~\eref{gge}.

@@

\paragraph*{Outline of the results.} 
In this paper, by combining exact Bethe ansatz and Monte Carlo techniques, we investigate 
the diagonal ensemble and the quench action approach in {\it finite} size integrable models. 
%
We consider the spin-$1/2$ XXZ spin-chain with $L$ sites, which is defined by the Hamiltonian 
%
\begin{equation}
\label{xxx-ham}
{\mathcal H}\equiv J\sum\limits_{i=1}^L\left[\frac{1}{2}(S_i^+S^-_{i+1} 
+S_i^{-}S_{i+1}^+)+\Delta S_i^zS_{i+1}^z-\frac{1}{4}\right].
\end{equation}
%
Here $S^{\pm}_i\equiv (\sigma_i^x\pm i\sigma_i^y)/2$ are spin operators acting on the 
site $i$, $S_i^z\equiv\sigma_i^z/2$, and $\sigma^{x,y,z}_i$ the Pauli matrices. We 
fix $J=1$ in~\eref{xxx-ham} and use periodic boundary conditions, identifying sites 
$L+1$ and $1$. The total magnetization $S_{T}^z\equiv\sum_iS_i^z=L/2-M$, with $M$ number 
of down spins (particles), commutes with~\eref{xxx-ham}, and it is used to label its 
eigenstates. The $XXZ$ chain is Bethe ansatz solvable~\cite{bethe-1931,kor-book} and its eigenstates 
are in one-to-one correspondence with the solutions of the so-called Bethe equations 
(see~\ref{sec:1.2}). 
%
The non-equilibrium quench dynamics of the XXZ spin chain has been investigated 
numerically intensively \cite{dmcf-06,bar-09,krs-12,cbp-13,cce-15,fagotti-2014}, the GGE with all the 
local charges has been explicitly constructed \cite{fagotti-2014,pozsgay-2013,fagotti-2013b}, 
and the quench action solution has been provided for some initial states~\cite{pozsgay-2014A,wouters-2014A}.

For concreteness reasons, we focus on the isotropic Heisenberg point ($XXX$), i.e. 
Hamiltonian~\eref{xxx-ham} with $\Delta=1$, but we stress that our approach applies to arbitrary 
$\Delta$, and  also to arbitrary integrable models. We consider the quenches from the zero-momentum 
N\'eel state $|N\rangle$ and the Majumdar-Ghosh state $|MG\rangle$, which are defined as  
%
\begin{equation}
\label{psi0}
|N\rangle\equiv\Big(\frac{|N_1\rangle
+|N_2\rangle}{\sqrt{2}}\Big), \qquad 
|MG\rangle\equiv \Big(\frac{|\uparrow\downarrow\rangle-|\downarrow\uparrow\rangle}
{\sqrt{2}}\Big)^{\otimes L/2}. 
\end{equation}
%
Here $|N_1\rangle\equiv |\uparrow\downarrow\rangle^{\otimes L/2}$, and $|N_2\rangle
\equiv |\downarrow\uparrow\rangle^{\otimes L/2}$, with $\otimes$ denoting the tensor 
product. Note that both $|N\rangle$ and $|MG\rangle$ are invariant under one site 
translations. Our study relies on the analytical knowledge of the overlaps between 
the N\'eel and Majumdar-Ghosh state and the eigenstates of~\eref{xxx-ham}~\cite{
brockmann-2014,brockmann-2014b,brockmann-2014c,pozsgay-2014a,piroli-2014,mazza-2015}.

@@

We also develop a Monte Carlo scheme, which is a generalization of 
the approach presented in Ref.~\cite{alba-2015} to simulate the Generalize 
Gibbs Ensemble. The method is based on the knowledge of the N\'eel overlaps 
$\langle\lambda|N\rangle$ and on the knowledge of the Hilbert space structure 
of the $XXX$ chain in the Bethe ansatz formalism. The approach allows us to 
simulate effectively chain with $L\lesssim 60$, although larger systems 
sizes can in principle be reached.

Strikingly, after the reweighting, although for small chains violations of 
the conserved quantities sum rules are present, these violations vanish in the 
thermodynamic limit and the sum rules are restored. This implies that the only 
effect of the Hilbert space truncation is to introduce finite size scaling corrections. 
In the quench action language, this means that the eigenstates corresponding to non 
zero-momentum strings contain enough physical information about the saddle point. 
This is numerically confirmed by extracting the so-called saddle point 
root distributions, which in the Bethe ansatz language fully characterize 
the diagonal ensemble averages in the thermodynamic limit. In the numerical  
 approach these are obtained from the histograms of the Bethe ansatz 
solutions sampled during the Monte Carlo. Apart from finite size scaling 
corrections, we observe striking agreement with analytical results, at least 
for the first few root distributions.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Parity-invariant eigenstates with non-zero N\'eel and Majumdar-Ghosh overlap: 
counting and string content}
\label{app-1}

Here we provide some exact combinatorial formulas for the total number of 
parity-invariant eigenstates of the $XXX$ chain in the sector with $L/2$ 
particles. These are the only eigenstates having, in principle, non-zero 
overlap with the N\'eel state and the Majumdar-Ghosh state~\cite{brockmann-2014}. 
Parity-invariant eigenstates correspond to solutions of the Bethe equations 
containing only pairs of rapidities with opposite sign. In turn, these  
eigenstates are in one-to-one correspondence with parity-invariant BGT quantum number 
configurations.

@@

We anticipate that the total number of parity-invariant eigenstates $Z_{Neel}$ 
for a chain of length $L$ is given as 
%
\begin{equation}
\label{N-count}
Z_{Neel}=2^{\frac{L}{2}-1}+\frac{1}{2}B\Big(\frac{L}{2},\frac{L}{4}\Big)+1. 
\end{equation}
%
On the other hand, after excluding the zero-momentum strings one obtains 
%
\begin{equation}
\label{N-count-nz}
\widetilde Z_{Neel}=B\Big(\frac{L}{2},\frac{L}{4}\Big). 
\end{equation}
%
Note that~\eref{N-count} is only an upper bound for the number of eigenstates 
with non-zero N\'eel overlap, while~\eref{N-count-nz} is exact. Before 
proceeding, we should stress that since the N\'eel state is not $SU(2)$ invariant, 
eigenstates with non-zero N\'eel overlap can contain infinite rapidities. Thus, 
one has to consider all the possible sectors with $L/2=\ell+N_{\infty}$, and $\ell$ 
($N_\infty$) the number of finite (infinite) rapidities. Notice that this 
is different for the Majumdar-Ghosh state, for which only the parity-invariant 
eigenstates in the sector with $\ell=L/2$ have to be considered (see below).

@@

Let us first consider the eigenstate sector with fixed number of finite rapidities 
$\ell$, the remaining $L/2-\ell$ ones being infinite (see section~\ref{sec:1.2}). 
Let us denote the associate string content as ${\mathcal S}=\{s_1,s_2,\dots,
s_{\ell}\}$. Here $s_n$ is the number of $n$-strings, with the constraint 
$\sum_k ks_k=\ell$. It is straightforward to check that the total number of 
parity-invariant quantum number pairs ${\mathcal N}_n(L,{\mathcal S})$ in the 
$n$-string sector is given as 
%
\begin{equation}
\label{NnLS}
{\mathcal N}_n(L,{\mathcal S})=\Big\lfloor\frac{L}{2}-\frac{1}{2}
\sum_{m=1}^{\ell}t_{nm}s_m\Big\rfloor,
\end{equation}
%
where $t_{nm}\equiv 2\textrm{Min}(n,m)-\delta_{n,m}$. Thus, the number of parity-invariant 
eigenstates of the $XXX$ chain ${\mathcal N}(L,{\mathcal S})$ compatible with string content 
${\mathcal S}$ is obtained by choosing in all the possible ways the associated parity-invariant 
quantum number pairs as     
%
\begin{equation}
\label{NLS}
{\mathcal N}(L,{\mathcal S})=\prod_{m=1}^{\ell} B\left({\mathcal N}_m,\left\lfloor
\frac{s_m}{2}\right\rfloor\right).
\end{equation}
%
Here the product is because each string sector is treated independently, while the 
factor $1/2$ in $s_m/2$ is because since all quantum numbers are organized in pairs, 
only half of them have to be specified. Note that in each $n$-string sector only one 
zero momentum (i.e., zero quantum number) string is allowed, due to the fact that 
repeated solutions of the BGT equation are discarded. Moreover, from~\eref{NnLS} one has 
that $s_m$ is odd (even) only if this zero momentum string is (not) present.

We now proceed to consider the string configurations with fixed particle number $\ell\le 
L/2$ and fixed number of strings $1\le q\le\ell$. Note that due to parity invariance 
$\ell$ must be even. 
Also, in determining $q$ strings of different length are treated equally, i.e., 
$q=\sum_m s_m$. For a given fixed pair 
$\ell,q$ the total number of allowed quantum number configurations by definition is given 
as 
%
\begin{equation}
\label{NLlq}
{\mathcal N}'(L,\ell,q)=\sum\limits_{\{\{s_m\}\,:\, \sum m s_m=\ell, \sum s_m=q\}}
{\mathcal N}(L,{\mathcal S}),
\end{equation}
%
where the sum is over the string content $\{s_m\}_{m=1}^\ell$ compatible with the constraints 
$\sum_m s_m=q$ and $\sum_m m s_m=\ell$. The strategy now is to write a recursive relation 
in both $\ell,q$ for ${\mathcal N}'(L,\ell,q)$. It is useful to consider a shifted string 
content ${\mathcal S}'$ defined as  
%
\begin{equation}
{\mathcal S}'\equiv \{s_{m+1}\}\quad\textrm{with}\, s_m\in{\mathcal S},\,\forall m.
\end{equation}
%
Using the definition of $t_{ij}$, it is straightforward to derive that  
%
\begin{equation}
t_{ij}=t_{i-1,j-1}+2,
\end{equation}
%
which implies that ${\mathcal N}_n(L,{\mathcal S})$ (see~\eref{NnLS}) satisfies the 
recursive equation 
%
\begin{equation}
\label{inter}
{\mathcal N}_n(L,{\mathcal S})={\mathcal N}_{n-1}(L-2q,{\mathcal S}'). 
\end{equation}
%
After substituting~\eref{inter} in~\eref{NLS} one obtains 
%
\begin{equation}
\label{NLSr}
{\mathcal N}(L,{\mathcal S})=B\Big({\mathcal N}_1(L,{\mathcal S}),\Big\lfloor 
\frac{s_1}{2}\Big\rfloor\Big){\mathcal N}(L-2q,{\mathcal S}'). 
\end{equation}
%
Finally, using~\eref{NLSr} in~\eref{NLlq}, one obtains a recursive relation for 
${\mathcal N}'(L,\ell,q)$ as 
%
\begin{equation}
\label{NpLlq}
{\mathcal N}'(L,\ell,q)=\sum_{s=0}^{q-1}B\Big(\frac{L}{2}-q+\Big\lfloor
\frac{s}{2}\Big\rfloor,\left\lfloor\frac{s}{2}\right\rfloor\Big){\mathcal N}'
\left(L-2q,\ell-q,q-s\right), 
\end{equation}
%
with the condition that for $\ell=q$ one has 
%
\begin{equation}
{\mathcal N}'(L,q,q)=B\Big(\Big\lfloor\frac{L-q}{2}\Big\rfloor,\Big\lfloor
\frac{q}{2} \Big\rfloor\Big).
\end{equation}
%
This is obtained by observing that if $\ell=q$ only $1$-strings are allowed 
and~\eref{NnLS} gives ${\mathcal N}_n(L,{\mathcal S})=\lfloor (L-q)/2\rfloor$. 
It is straightforward to check that to satisfy~\eref{NpLlq} for $q$ even one has 
to choose 
%
\begin{equation}
\label{inter1}
{\mathcal N}'(L,\ell,q)=\frac{q}{\ell}B\Big(\frac{L-\ell}{2},\frac{q}{2}\Big)
B\Big(\frac{\ell}{2},\frac{q}{2}\Big),
\end{equation}
% 
Instead, for $q$ odd one has
%
\begin{equation}
\label{inter2}
{\mathcal N}'(L,\ell,q)=\frac{\ell-q+1}{\ell}B\Big(\frac{L-\ell}{2},\frac{q-1}{2}
\Big)B\Big(\frac{\ell}{2},\frac{q-1}{2}\Big).
\end{equation}
%
The number of eigenstates in the sector with $\ell$ particles having nonzero N\'eel 
overlap $Z'_{Neel}(L,\ell)$ is obtained from~\eref{inter1} and~\eref{inter2} by 
summing over all possible values of $q$ as 
%
\begin{equation}
\label{sum1}
Z'_{Neel}(L,\ell)=\sum\limits_{q=1}^\ell {\mathcal N}'(L,\ell,q).
\end{equation}
%
It is convenient to split the summation in~\eref{sum1} considering odd and even 
$q$ separately. For $q$ odd one obtains 
%
\begin{equation}
\sum\limits_{k=0}^{\ell/2-1} {\mathcal N}'(L,\ell,2k+1)=B\Big(\frac{L}{2}-1,
\frac{\ell}{2}-1\Big),
\end{equation}
%
while for $q$ even one has 
%
\begin{equation}
\sum\limits_{k=0}^{\ell/2} {\mathcal N}'(L,\ell,2k)=B\Big(\frac{L}{2}-1,
\frac{\ell}{2}\Big). 
\end{equation}
%
Putting everything together one obtains 
%
\begin{equation}
\label{N-count-p}
Z'_{Neel}(L,\ell)=B\Big(\frac{L}{2}-1,
\frac{\ell}{2}-1\Big)+B\Big(\frac{L}{2}-1,
\frac{\ell}{2}\Big). 
\end{equation}
%
The total number of eigenstates with nonzero N\'eel overlap $Z_{Neel}(L)$ 
(cf.~\eref{N-count}) is obtained from~\eref{N-count-p} by summing over the allowed 
values of $\ell=2k$ with $k=0,1,\dots,L/2$. Note that the sum is over $\ell$ even 
due to the parity invariance.

@@

\bibitem{ga-15}
G. Goldstein and N. Andrei, Phys. Rev. B \textbf{92}, 155103 (2015).
